The cardinality-constrained college application problem: The nestedness property and implications for student welfare

### 2022-07-28 paper_alma/body.tex
risk aversion discussion

The distributional heuristic does not specify the boundary between target, safety, and reach schools. However, regardless of how this boundary is specified, we argue that the distributional heuristic is limited by the fact that it does not respond adaptively to the application budget $h$. In a market where $f_j$ and $t_j$ are inversely correlated, we find that the first schools to enter the optimal portfolio in the greedy algorithm are safety schools, and the allocation shifts to incorporate more target and reach schools as the application budget $h$ grows. This phenomenon, which we call \emph{optimal risk elasticity,} implies that any heuristic that targets a \emph{static} risk distribution becomes excessively risk averse for students with a sufficiently large application budget. Students with a large application budget can tolerate, and indeed benefit, from taking larger risks than those without.

To see why large values of $h$ favor a riskier portfolio allocation, consider two schools $i$ and $j$ having $t_i < t_j$ and $f_i t_i = f_j t_j = 1$. Thus, the expected utility of applying \emph{individually} to either school is identical, and a risk-neutral student is indifferent between the safety school $i$ and the reach school $j$. Now suppose that the student has resolved to apply to another school, $k$, and the question is whether to apply to  $\mathcal{X} = \{i, k\}$ or $\{j, k\}$. There are three cases:

@@

Regression studies of college applicant behavior have shown that low-income students tend to apply to a few safety schools, whereas their wealthy peers tend to apply to more schools and take bigger risks \cite{hoxbyandavery2012}. One interpretation of these findings is as part of a larger pattern of greater risk aversion in those from disadvantaged backgrounds, a finding that has been noted in other areas of psychological research \cite{hartlaubandschneider2012,wattsetal2018,vanhuizenandalessie2019}. Optimal risk elasticity, however, suggests a rational basis for low-income students' apparent risk aversion in the college application process: Applying to a few safety schools is often the \emph{optimal} strategy for a risk-neutral applicant with a small budget. % This analysis echoes arguments such as (cite) that suggest that risk aversion in the marshmallow test reflects students rational expectation of being rewarded for good behavior.

Optimal risk elasticity also differentiates the college application problem from classical portfolio optimization models in the Markowitz \cite{markowitz1952} tradition, which feature risk aversion as an exogeneous \emph{parameter.} In the Markowitz model, the optimal portfolio is expressed as a proportion of the total budget, and this allocation is constant for budgets of all sizes. Therefore, the value of a rational investor's risk parameter can be determined merely by locating her portfolio allocation along the efficient frontier, even if her total investment budget is unknown. In college application, on the other hand, variation in students' risk allocation arises from a combination of variance in underlying preferences and variance in students' application budgets. For this reason, we propose that future econometric studies of risk aversion in college application control for the size of the application portfolio in comparing students' allocation across reach, target, and safety schools.

@@

Diminishing returns are an intrinsic feature of submodular set functions. The college application problem is no exception, as the following theorem states.

@@

\begin{proof}
Applying Theorem \ref{nestedapplication}, we write $\mathcal{X}_h = \mathcal{X}_{h-1} \cup\{j\}$ and $\mathcal{X}_{h+1} = \mathcal{X}_{h-1} \cup\{j, k\}$. By optimality, $v(\mathcal{X}_h) - v(\mathcal{X}_{h-1}) \geq v(\mathcal{X}_{h-1}\cup\{k\}) - v(\mathcal{X}_{h-1})$. By submodularity and nestedness, $v(\mathcal{X}_{h-1}\cup\{k\}) - v(\mathcal{X}_{h-1}) \geq  v(\mathcal{X}_{h}\cup\{k\}) - v(\mathcal{X}_{h}) = v(\mathcal{X}_{h+1}) - v(\mathcal{X}_{h})$.
\end{proof}

\noindent (An elementary proof appears in \cite{kapur2022}.) It follows that when $\mathcal{X}_h$ is the optimal $h$-portfolio, for fixed values of $f$ and $t$, $v(\mathcal{X}_h) = t_0 + O(h)$. The following example establishes the tightness of this bound.

@@

In Example \ref{tightexampleforconcavity}, the ability of the student to achieve a valuation linear in $h$ hinges on the fact that the admissions probabilities $f_j$ are very small; when some schools have $f_j$ close to one, the valuation function tends to flatten out as these schools are incorporated into the optimal portfolio. This means that students with many reach schools in their portfolio benefit more from the opportunity to apply to an additional college than students with many safety schools.

@@

% explicit risk aversion parameter

>>>>

The distributional heuristic does not specify the boundary between target, safety, and reach schools in numerical terms. However, regardless of how this boundary is specified, we argue that the distributional heuristic is limited by the fact that it does not respond adaptively to the application budget $h$. In a market where $f_j$ and $t_j$ are inversely correlated, the first schools to enter the optimal portfolio in the greedy algorithm are safety schools, and the allocation shifts to incorporate more target and reach schools as the application budget $h$ grows. This phenomenon, which we call \emph{optimal risk elasticity,} implies that any heuristic that targets a \emph{static} risk distribution becomes excessively risk averse for students with a sufficiently capacious application budget. Students with a large budget can tolerate, and indeed benefit, from taking bigger risks than those without.

To see why large values of $h$ favor a riskier portfolio allocation, consider two schools $i$ and $j$ having $t_i < t_j$ and $f_i t_i = f_j t_j = 1$. That is, the expected utility of applying \emph{individually} to either school is identical, and a risk-neutral student is indifferent between the safety school $i$ and the reach school $j$. Now suppose that the student has resolved to apply to another school, $k$, and the question is whether to apply to  $\mathcal{X} = \{i, k\}$ or $\{j, k\}$. There are three cases:

@@

\subsection{Implications of optimal risk elasticity for econometric research} \label{subsectionRiskAversionEconometrics}

Our formulation of the college application problem has presumed a risk-neutral student who maximizes her portfolio's expected value without regard to its variance. Is this model complex enough to capture the psychology of real college applicants? Perhaps not: Regression studies of college applicant behavior have shown that low-income students tend to apply to a few safety schools, whereas their wealthy peers tend to apply to more schools and take bigger risks \cite{hoxbyandavery2012}. One interpretation of these findings is as part of a larger pattern of greater risk aversion in those from disadvantaged backgrounds, a finding that has been noted in other areas of psychological research \cite{hartlaubandschneider2012,wattsetal2018,vanhuizenandalessie2019}.

Risk aversion almost certainly factors into real students’ decisionmaking; however, optimal risk elasticity implies that students may \emph{appear} to differ systematically in their risk preferences as an artifact of underlying differences in application resources. For a student with no risk aversion but a small application budget, applying to a few safety schools is often the \emph{rational} strategy. Thus, to the extent that low-income students have smaller application budgets than their wealthy peers, it may be incorrect to attribute variance in students' allocation across reach, target, and safety schools solely or primarily to psychological risk aversion. We propose that future regression studies of applicant behavior control for the number of schools in the application portfolio when comparing students’ risk allocation across demographic categories.

Optimal risk elasticity distinguishes the college application problem from classical portfolio optimization models in the Markowitz \cite{markowitz1952} tradition, which feature an exogeneous risk-aversion parameter. In the Markowitz model, the optimal portfolio is expressed as a proportion of the total budget, and this allocation is constant for budgets of all sizes. Therefore, if an investment adviser knows a client’s risk parameter, then she can recommend an optimal portfolio allocation without asking the client’s total investment budget. In college application, on the other hand, even among students with \emph{no} risk aversion and identical input data $f$ and $t$, the proportion of the portfolio allocated to safe and risky assets depends on the budget $h$. Allowing for variation in risk aversion and the market parameters only complicates this picture further.

@@

Diminishing returns to the size of $\mathcal{X}$ are an intrinsic feature of submodular set functions. The college application problem is no exception, as the following theorem states.

@@

\begin{proof}
Applying Theorem \ref{nestedapplication}, we write $\mathcal{X}_h = \mathcal{X}_{h-1} \cup\{j\}$ and $\mathcal{X}_{h+1} = \mathcal{X}_{h-1} \cup\{j, k\}$. By optimality, $v(\mathcal{X}_h) - v(\mathcal{X}_{h-1}) \geq v(\mathcal{X}_{h-1}\cup\{k\}) - v(\mathcal{X}_{h-1})$. By submodularity, $v(\mathcal{X}_{h-1}\cup\{k\}) - v(\mathcal{X}_{h-1}) \geq  v(\mathcal{X}_{h}\cup\{k\}) - v(\mathcal{X}_{h}) = v(\mathcal{X}_{h+1}) - v(\mathcal{X}_{h})$.
\end{proof}

\noindent (An elementary proof appears in \cite{kapur2022}.) It follows that when $\mathcal{X}_h$ is the optimal $h$-portfolio for fixed values of $f$ and $t$, $v(\mathcal{X}_h) = t_0 + O(h)$. The following example establishes the tightness of this bound.

@@

In Example \ref{tightexampleforconcavity}, the ability of the student to achieve a valuation linear in $h$ hinges on the fact that the admissions probabilities $f_j$ are very small; when some schools have $f_j$ close to one, the valuation function tends to ``flatten out'' as these schools are incorporated into the optimal portfolio. This means that, ceteris paribus, students with many reach schools in their portfolio benefit more from the opportunity to apply to an additional college than students with many safety schools.

@@

A potential application of our model is an econometric study that estimates students' perceptions of college quality ($t_j$) from their observed application behavior ($\mathcal{X}$) under the assumption of utility maximization. As discussed in section \ref{subsectionRiskAversionEconometrics}, real-world applicant behavior draws on a combination of utility-maximizing and risk-averse strategies. Therefore, adding a variance penalty of the following form to the objective function could improve our model's ability to capture real student behavior:
\begin{equation*}
\tilde v(\mathcal{X})
= \operatorname{E}\Bigl[\max\bigr\{t_0, \max\{t_j Z_j : j \in \mathcal{X}\}\bigr\}\Bigr] 
- \gamma \operatorname{Var}\Bigl[\max\bigr\{t_0, \max\{t_j Z_j : j \in \mathcal{X}\}\bigr\}\Bigr]
\end{equation*}
($\gamma > 0$ is the risk-aversion parameter.) Our study has not considered the portfolio optimization problem associated with this more complicated objective, let alone the econometric task of estimating its parameters from applicant behavior.
